---
title: "Community Cohort Tool"
author: "Chicago Metropolitan Agency for Planning"
date: "`r format(Sys.time(), '%x')`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
#install.packages("tidyverse", "readxl", "ggplot2", "sf", "tmap")
library(tidyverse)
library(readxl)
library(ggplot2)
library(sf)
library(tmap)
library(tmaptools)
```

This tool is used by the [Chicago Metropolitan Agency for Planning (CMAP)](https://www.cmap.illinois.gov) to group the 284 municipalities and 77 Chicago Community Areas in the region into Community Cohorts, which are then used to determine local contribution levels for most [Local Technical Assistance](https://www.cmap.illinois.gov/programs/LTA) projects. The source code and data is hosted in a [GitHub repository](https://github.com/CMAP-REPOS/community-cohort-tool).

The baseline local contribution requirement for LTA projects is 20% of the project costs. Reduced contributions are granted to higher-need communities, based on their Community Cohort:

* **Community Cohort 1:** 100% of baseline local contribution requirement (no reduction)
* **Community Cohort 2:** 75% of baseline local contribution requirement
* **Community Cohort 3:** 50% of baseline local contribution requirement
* **Community Cohort 4:** 25% of baseline local contribution requirement


# Load input factors, weights and cohort thresholds

This tool relies on a single Excel file -- located at **input/community_cohort_inputs.xlsx** -- for all of its data and parameters. The file contains 4 sheets, which can be modified as needed to alter the tool:

* `FACTORS_MUNI` -- a table of potential factors to include in the cohort scoring for each municipality
* `FACTORS_CCA` -- a table of potential factors to include in the cohort scoring for each Chicago Community Area (CCA)
* `WEIGHTS` -- a table of weights assigned to individual factors present in both `FACTORS_MUNI` and `FACTORS_CCA`; factors omitted from this table (or given a weight of 0) will be excluded from the cohort scoring
* `COHORTS` -- a table defining the thresholds for assigning communities to cohorts, based on their overall/weighted scores


### Potential factors (municipalities)

Load a table of potential factors for each of the 284 municipalities in the CMAP region from the **FACTORS_MUNI** sheet in **input/community_cohort_inputs.xlsx**. This table may include data that are not used in cohort scoring, depending on the contents of the **WEIGHTS** sheet (see below).

```{r echo=FALSE, message=FALSE, warning=FALSE}
FACTORS_MUNI <- read_xlsx("input/community_cohort_inputs.xlsx", sheet="FACTORS_MUNI")
FACTORS_MUNI
```


### Potential factors (Chicago Community Areas)

Load a table of potential factors for each of the 77 Chicago Community Areas (CCAs) within the City of Chicago from the **FACTORS_CCA** sheet in **input/community_cohort_inputs.xlsx**. This table may include data that are not used in cohort scoring, depending on the contents of the **WEIGHTS** sheet (see below).

```{r echo=FALSE, message=FALSE, warning=FALSE}
FACTORS_CCA <- read_xlsx("input/community_cohort_inputs.xlsx", sheet="FACTORS_CCA")
FACTORS_CCA
```


### Weights

Load a table of weights for the factors from the **WEIGHTS** sheet in **input/community_cohort_inputs.xlsx**. Factors with non-zero weights will be included in the scoring. The larger the *absolute magnitude* of a factor, the more influence it will have on the overall score. Negative weights should be specified for factors where lower values correspond to lower need (i.e. the lowest values will receive the highest scores).

```{r echo=FALSE, message=FALSE, warning=FALSE}
WEIGHTS <- read_xlsx("input/community_cohort_inputs.xlsx", sheet="WEIGHTS")
WEIGHTS
```


### Cohort thresholds

Load a table of cohort thresholds from the **COHORTS** sheet in **input/community_cohort_inputs.xlsx**. These will be the maximum overall scores (from a possible range of 0-100) that will be assigned to a particular cohort.

```{r echo=FALSE, message=FALSE, warning=FALSE}
COHORTS <- read_xlsx("input/community_cohort_inputs.xlsx", sheet="COHORTS")
COHORTS$COHORT <- as.character(COHORTS$COHORT)
COHORTS
```


# Calculate factor-specific scoring thresholds

For each factor included in the `WEIGHTS` table, calculate the median and standard deviation from the **FACTORS_MUNI** table. Use these with specific z-scores to assign municpalities a score from 1-10, such that each group will roughly represent an equal number of municipalities (assuming the factor approximates a normal distribution). 

```{r echo=FALSE, message=FALSE, warning=FALSE}
WEIGHTS$MED <- unlist(summarize_all(FACTORS_MUNI[, WEIGHTS$FACTOR_NAME], median)[1,])
WEIGHTS$SD <- unlist(summarize_all(FACTORS_MUNI[, WEIGHTS$FACTOR_NAME], sd)[1,])

WEIGHTS <- WEIGHTS %>%
  mutate(
    CUT0 = -Inf,
    CUT1 = MED - SD * 1.2816,  # ~10th %ile (based on standard normal distribution)
    CUT2 = MED - SD * 0.8416,  # ~20th %ile
    CUT3 = MED - SD * 0.5244,  # ~30th %ile
    CUT4 = MED - SD * 0.2533,  # ~40th %ile
    CUT5 = MED,                # ~50th %ile
    CUT6 = MED + SD * 0.2533,  # ~60th %ile
    CUT7 = MED + SD * 0.5244,  # ~70th %ile
    CUT8 = MED + SD * 0.8416,  # ~80th %ile
    CUT9 = MED + SD * 1.2816,  # ~90th %ile
    CUT10 = Inf
  )
```

For factors that do not approximate a normal distribution, make manual adjustments to the group thresholds as necessary to obtain sensible scores.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Force equal intervals and midpoint of 0.5 for PCT_EDA_POP factor
WEIGHTS[WEIGHTS$FACTOR_NAME=="PCT_EDA_POP", paste0("CUT", 1:9)] <- seq(0.1, 0.9, 0.1)
WEIGHTS %>% select(FACTOR_NAME, WEIGHT, starts_with("CUT"), -CUT0, -CUT10)
```


# Calculate factor-specific scores

For each factor, use the calculated group thresholds to assign factor-specific scores to each municipality. Factors with negative weights will have their scores reversed so that the lowest values receive the highest scores. Inspect histograms of the factors, as well as the resultant scoring distributions

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.asp=0.5}
keep_cols_muni <- append(c("GEOID", "MUNI"), WEIGHTS$FACTOR_NAME)
keep_cols_cca <- append(c("CCA_ID", "CCA_NAME"), WEIGHTS$FACTOR_NAME)
FACTORS_MUNI <- FACTORS_MUNI[, keep_cols_muni]
FACTORS_CCA <- FACTORS_CCA[, keep_cols_cca]

score_cols <- c()
wt_score_cols <- c()

for (factor in WEIGHTS$FACTOR_NAME) {
  weight <- WEIGHTS[WEIGHTS$FACTOR_NAME==factor, "WEIGHT"][[1]]

  score_col <- paste0("SCORE_", factor)
  score_cols <- append(score_cols, score_col)

  wt_score_col <- paste0("WT_SCORE_", factor)
  wt_score_cols <- append(wt_score_cols, wt_score_col)

  cuts <- WEIGHTS[WEIGHTS$FACTOR_NAME==factor,] %>%
    select(starts_with("CUT"))
  groups <- c(1:10)
  FACTORS_MUNI[, score_col] <- cut(as.matrix(FACTORS_MUNI[, factor]), cuts, groups, labels=FALSE)
  FACTORS_CCA[, score_col] <- cut(as.matrix(FACTORS_CCA[, factor]), cuts, groups, labels=FALSE)
  if (weight < 0) {
    # Reverse score order for factors with negative weights
    FACTORS_MUNI[, score_col] <-  max(groups) + 1 - FACTORS_MUNI[, score_col]
    FACTORS_CCA[, score_col] <-  max(groups) + 1 - FACTORS_CCA[, score_col]
  }
  FACTORS_MUNI[, wt_score_col] <- FACTORS_MUNI[, score_col] * abs(weight)
  FACTORS_CCA[, wt_score_col] <- FACTORS_CCA[, score_col] * abs(weight)

  # Inspect score distribution
  print(ggplot(FACTORS_MUNI) +
          geom_histogram(aes(x=get(factor)), color="white", fill="skyblue", bins=50) +
          geom_vline(xintercept=cuts[[2]], color="maroon", linetype="dotted") +
          geom_vline(xintercept=cuts[[3]], color="maroon", linetype="dotdash") +
          geom_vline(xintercept=cuts[[4]], color="maroon", linetype="dashed") +
          geom_vline(xintercept=cuts[[5]], color="maroon", linetype="longdash") +
          geom_vline(xintercept=cuts[[6]], color="maroon", linetype="solid", size=1) +  # Median
          geom_vline(xintercept=cuts[[7]], color="maroon", linetype="longdash") +
          geom_vline(xintercept=cuts[[8]], color="maroon", linetype="dashed") +
          geom_vline(xintercept=cuts[[9]], color="maroon", linetype="dotdash") +
          geom_vline(xintercept=cuts[[10]], color="maroon", linetype="dotted") +
          labs(title="Distribution of factor values (with group breaks)",
               subtitle=factor,
               x=factor, y="Number of municipalities") +
          theme_classic())

  print(ggplot(FACTORS_MUNI) +
          geom_histogram(aes(x=get(score_col)), color="white", fill="skyblue", binwidth=1) +
          geom_hline(yintercept=28.4, color="maroon", linetype="dashed") +
          scale_x_continuous(limits=c(min(groups)-1, max(groups)+1), breaks=groups) +
          labs(title="Distribution of factor scores", subtitle=score_col,
               caption="Dashed line represents a perfect decile distribution of 28.4 municipalities per group",
               x=score_col, y="Number of municipalities") +
          theme_classic())
}
```

The same thresholds derived to assign scores to the municipalities will also be used for assigning scores to CCAs, which were not included in the charts above.


# Calculate overall score & cohorts

The overall score is calculated for each community by multiplying each factor-specific score by the absolute magnitude of that factor's weight, summing these products, and then scaling the results so that the minimum and maximum theoretical scores (i.e. 1's and 10's for every factor-specific score, respectively) would receive overall scores of 0 and 100, respectively.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.asp=0.7}
FACTORS_MUNI$SCORE_OVERALL <- rowSums(FACTORS_MUNI[, wt_score_cols])
FACTORS_CCA$SCORE_OVERALL <- rowSums(FACTORS_CCA[, wt_score_cols])

# Rescale from 0-100
min_wt_score <- sum(abs(WEIGHTS$WEIGHT)) * 1
max_wt_score <- sum(abs(WEIGHTS$WEIGHT)) * 10

FACTORS_MUNI <- FACTORS_MUNI %>%
  mutate(SCORE_OVERALL_SCALED = (SCORE_OVERALL - min_wt_score) / (max_wt_score - min_wt_score) * 100)
FACTORS_MUNI$COHORT <- cut(as.matrix(FACTORS_MUNI$SCORE_OVERALL_SCALED), c(-Inf, COHORTS$MAX_SCORE), COHORTS$COHORT)
FACTORS_MUNI <- FACTORS_MUNI %>%
  mutate(COHORT = fct_relevel(COHORT, sort))

ggplot(FACTORS_MUNI) +
  geom_histogram(aes(x=SCORE_OVERALL_SCALED, fill=COHORT), color="white", binwidth=5, center=2.5) +
  scale_x_continuous(limits=c(0, 100), breaks=seq(0, 100, 20)) +
  scale_fill_discrete(name="Assigned cohort") +
  labs(title="Distribution of overall scores (municipalities)", x="Overall score", y="Number of municipalities") +
  theme_classic()

FACTORS_CCA <- FACTORS_CCA %>%
  mutate(SCORE_OVERALL_SCALED = (SCORE_OVERALL - min_wt_score) / (max_wt_score - min_wt_score) * 100)
FACTORS_CCA$COHORT <- cut(as.matrix(FACTORS_CCA$SCORE_OVERALL_SCALED), c(-Inf, COHORTS$MAX_SCORE), COHORTS$COHORT)
FACTORS_CCA <- FACTORS_CCA %>%
  mutate(COHORT = fct_relevel(COHORT, sort))

chi_overall <- FACTORS_MUNI[FACTORS_MUNI$MUNI=="Chicago", "SCORE_OVERALL_SCALED"][[1]]
ggplot(FACTORS_CCA) +
  geom_histogram(aes(x=SCORE_OVERALL_SCALED, fill=COHORT), color="white", binwidth=5, center=2.5) +
  geom_vline(linetype="dashed", xintercept=chi_overall) +
  scale_x_continuous(limits=c(0, 100), breaks=seq(0, 100, 20)) +
  scale_fill_discrete(name="Assigned cohort") +
  labs(title="Distribution of overall scores (CCAs)",
       caption="Dashed line represents the overall score for the entire City of Chicago",
       x="Overall score", y="Number of CCAs") +
  theme_classic()
```


# Map the results

Create maps showing the final cohort assignment for each municipality and CCA.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.asp=1.3}
IL_E_NAD83 <- 26971  # EPSG projection ID: IL State Plane East (NAD83)
tmap_mode("plot")  # "plot" (static) or "view" (interactive)

cnty_geo <- st_read("input/cmap_county_boundaries.geojson", quiet=TRUE) %>%
  st_transform(IL_E_NAD83)

muni_geo <- st_read("input/cmap_munis.geojson", quiet=TRUE) %>%
  st_transform(IL_E_NAD83) %>%
  left_join(FACTORS_MUNI, by=c("GEOID_n"="GEOID"))

muni_labels <- muni_geo %>%
  filter(MUNI.x %in% c("Chicago", "Joliet", "Aurora", "Elgin", "Waukegan"))  # Label select munis

cca_geo <- st_read("input/chicago_ccas.geojson", quiet=TRUE) %>%
  st_transform(IL_E_NAD83) %>%
  left_join(FACTORS_CCA, by=c("CCA_NUM"="CCA_ID"))

tm_shape(muni_geo, bbox=bb(cnty_geo, ext=1.2)) +
  tm_polygons("COHORT", title="", palette="Reds", n=4, border.col="#ffffff",
              labels=c("1 (low need)", "2 (moderate need)", "3 (high need)", "4 (very high need)")) +
tm_shape(cnty_geo) +
  tm_lines(col="#888888", lwd=2) +
tm_shape(muni_labels) +
  tm_text("MUNI.x", size=0.7, col="#000000", fontface="bold") +
tm_legend(legend.position=c("left", "bottom")) +
tm_layout(title="Assigned cohorts (municipalities)", frame=FALSE)

tm_shape(cca_geo, bbox=bb(cca_geo, ext=1.2)) +
  tm_polygons("COHORT", title="", palette="Reds", n=4, border.col="#ffffff",
              labels=c("1 (low need)", "2 (moderate need)", "3 (high need)", "4 (very high need)")) +
tm_legend(legend.position=c("left", "bottom")) +
tm_layout(title="Assigned cohorts (CCAs)", frame=FALSE)
```


# Write output files

Write overall scores and assigned cohorts for each municipality and CCA in the following files:

### output/cohort_assignments_muni.csv

```{r echo=FALSE, message=FALSE, warning=FALSE}
OUT_DATA_MUNI <- FACTORS_MUNI %>%
  rename(WEIGHTED_SCORE = SCORE_OVERALL_SCALED) %>%
  select(GEOID, MUNI, COHORT, WEIGHTED_SCORE, starts_with("SCORE_")) %>%
  select(-SCORE_OVERALL)
OUT_DATA_MUNI

write_csv(OUT_DATA_MUNI, "output/cohort_assignments_muni.csv")
```

### output/cohort_assignments_cca.csv

```{r echo=FALSE, message=FALSE, warning=FALSE}
OUT_DATA_CCA <- FACTORS_CCA %>%
  rename(WEIGHTED_SCORE = SCORE_OVERALL_SCALED) %>%
  select(CCA_ID, CCA_NAME, COHORT, WEIGHTED_SCORE, starts_with("SCORE_")) %>%
  select(-SCORE_OVERALL)
OUT_DATA_CCA

write_csv(OUT_DATA_CCA, "output/cohort_assignments_cca.csv")
```


# Compare scores/cohorts against previous methodology

View a table of municipalities whose assigned cohort is different than it was with the previous methodology. Get some descriptive statistics about how different the cohort assignments are overall.

```{r echo=FALSE, message=FALSE, warning=FALSE}
PREV_SCORES <- read_csv("input/original_scores_fy20.csv", col_types=cols(COHORT=col_character())) %>%
  rename(
    SCORE_PREV = FINAL_SCORE,
    COHORT_PREV = COHORT
  )

COMPARE <- FACTORS_MUNI %>%
  left_join(PREV_SCORES, by="MUNI") %>%
  mutate(
    COHORT_CHG = as.numeric(COHORT) - as.numeric(COHORT_PREV),
    POP = exp(ln_POP)
  )

CHANGED_MUNIS <- COMPARE[COMPARE$COHORT != COMPARE$COHORT_PREV, c("MUNI", "COHORT", "COHORT_PREV", "POP")]
CHANGED_MUNIS

cat("Correlation (R-squared):", cor(COMPARE$SCORE_PREV, COMPARE$SCORE_OVERALL_SCALED))

lm_old_new <- lm(SCORE_OVERALL_SCALED ~ SCORE_PREV, data=COMPARE)
cat("Linear trend: SCORE_new = ", lm_old_new$coefficients[1], " + ", lm_old_new$coefficients[2], "*SCORE_old", sep="")

library(modelr)
cat("Root-mean-square error (RMSE):", rmse(lm_old_new, COMPARE))
```

Plot some charts comparing previous scores/cohorts to the current ones.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.asp=0.7}
ggplot(COMPARE) +
  geom_point(aes(x=SCORE_PREV, y=SCORE_OVERALL_SCALED, color=COHORT_PREV), alpha=0.6) +
  geom_abline(intercept=0, slope=1, color="gray", linetype="dashed") +
  geom_abline(intercept=lm_old_new$coefficients[1], slope=lm_old_new$coefficients[2]) +
  scale_x_continuous(limits=c(0, 100), breaks=seq(0, 100, 10)) +
  scale_y_continuous(limits=c(0, 100), breaks=seq(0, 100, 10)) +
  scale_color_discrete(name="Previous cohort") +
  labs(title="Comparison of updated scores vs. previous scores", x="Previous score", y="Updated score") +
  theme_classic()

ggplot(COMPARE) +
  geom_count(aes(x=COHORT_PREV, y=COHORT), color="maroon") +
  scale_size_area(max_size = 20) +
  labs(title="Communities by previous and updated cohort", x="Previous cohort", y="Updated cohort")

ggplot(COMPARE) +
  geom_histogram(aes(x=COHORT_PREV, fill="Previous"), stat="count", width=0.4, position=position_nudge(x=-0.2)) +
  geom_histogram(aes(x=COHORT, fill="Updated"), stat="count", width=0.4, position=position_nudge(x=0.2)) +
  labs(title="Comparison of updated cohorts vs. previous cohorts",
       x="Cohort", y="Number of municipalities") +
  theme_classic() +
  theme(legend.title=element_blank())

COHORT_POP <- COMPARE %>%
  select(COHORT, POP) %>%
  group_by(COHORT) %>%
  summarize(POP = sum(POP))

COHORT_POP_PREV <- COMPARE %>%
  select(COHORT_PREV, POP) %>%
  group_by(COHORT_PREV) %>%
  summarize(POP = sum(POP))

COHORT_POP_CHG <- COHORT_POP %>%
  left_join(COHORT_POP_PREV, by=c("COHORT" = "COHORT_PREV"), suffix=c("", "_PREV"))

ggplot(COHORT_POP_CHG) +
  geom_col(aes(x=COHORT, y=POP_PREV, fill="Previous"), width=0.4, position=position_nudge(x=-0.2)) +
  geom_col(aes(x=COHORT, y=POP, fill="Updated"), width=0.4, position=position_nudge(x=0.2)) +
  labs(title="Comparison of population in updated cohorts vs. previous cohorts",
       x="Cohort", y="Total population") +
  theme_classic() +
  theme(legend.title=element_blank())
```

Create a map highlighting where the specific differences are.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.asp=1.3}
muni_geo <- st_read("input/cmap_munis.geojson", quiet=TRUE) %>%
  st_transform(IL_E_NAD83) %>%
  left_join(COMPARE, by=c("GEOID_n"="GEOID"))

tm_shape(muni_geo, bbox=bb(cnty_geo, ext=1.2)) +
  tm_polygons("COHORT_CHG", title="", palette="-PuOr", contrast=c(0,1), n=7, border.col="#ffffff",
              midpoint=NA, style="fixed", breaks=c(-3,-2,-1,0,1,2,3,4),
              labels=c("-3 (lower need)", "-2", "-1", "+0 (no change)", "+1", "+2", "+3 (higher need)")) +
tm_shape(cnty_geo) +
  tm_lines(col="#888888", lwd=2) +
tm_shape(muni_labels) +
  tm_text("MUNI.x", size=0.7, col="#000000", fontface="bold") +
tm_legend(legend.position=c("left", "bottom")) +
tm_layout(title="Change in cohort (previous to updated)", frame=FALSE)
```

